From the function-sheaf dictionary to quasicharacters of $p$-adic tori

### 2015-06-01 dictionary.tex
Now the conflicts in my office are really resolved.

\begin{abstract}
We consider the category of character sheaves on smooth commutative group schemes $G$ over finite fields $k$ and expand the scope of the function-sheaf dictionary from connected commutative algebraic groups to this setting.
We show that the group of isomorphism classes of character sheaves is an extension of the group of characters of $G(k)$ by a cohomology group determined by the component group scheme of $G$.
As an application, we find a category of sheaves for which isomorphism classes correspond to quasi\-characters in the case of algebraic tori, abelian varieties and unipotent wound groups over local fields. 
In this geometrization of quasicharacters, our sheaves are local systems on
Greenberg transforms of locally finite type N\'eron models.
\end{abstract}

@@

This paper extends Deligne's function-sheaf dictionary from connected commutative algebraic groups to smooth commutative group schemes over finite fields and then provides an application to the geometrization and categorification of quasicharacters of $p$-adic tori.

@@

We begin the paper by returning to the case of connected commutative algebraic groups $G$ and thus revisiting Deligne's function-sheaf dictionary.
We consider character sheaves that arise via base change to $\bFq$ from local systems on $G$ (\S\ref{ssec:descentG}) and
those that appear in a pushforward from a constant sheaf along a discrete isogeny $H \to G$ (\S\ref{ssec:discrete_isogenies}).  
While these constructions make sense even for non-connected $G$, in the connected case we show that every character sheaf can be described in both of these ways (\S\ref{ssec:connected}).  
We use this fact to prove that $\TrFrob{G}$ is an isomorphism for connected commutative algebraic groups $G$
and to determine the automorphism groups of character sheaves on such $G$. 
These results are all well know.

Next, we consider character sheaves on \'etale commutative group schemes $G$ over $\Fq$ (\S\ref{sec:etale});
\'etale group schemes form a counterpoint to connected algebraic groups, since the component group of any smooth group scheme is an \'etale group scheme.
Our key tool for understanding the trace of Frobenius in the \'etale case 
%are a reinterpretation of $\CS(G)$ in terms of stalks (\S\ref{ssec:stalks}) and 
is a relationship with the Hochschild-Serre spectral sequence (\S\ref{ssec:E}) for $\Weil{} \ltimes \bG$, where $\Weil{} \subset \Gal(\Fq/\Fq)$ is the Weil group for $\Fq$.
We define (\S\ref{ssec:S}) an isomorphism $S_G$ from $\CSiso{G}$ to the second cohomology of the total space of the spectral sequence
\[
E_2^{p,q} \ceq \Hh^p(\Weil{}, \Hh^q(\bG, \EEx)) \Rightarrow \Hh^{p+q}(\Weil{} \ltimes \bG, \EEx).
\]
Paired with the short exact sequence
\[
  0 \to \Hh^0(\Weil{},\Hh^2({\bar G},\EEx)) \to \Hh^2(E^\bullet_G) \to \Hh^1(\Weil{},\Hh^1({\bar G},\EEx)) \to 0
\]
arising from the spectral sequence, the isomorphism $S_G$ allows us to show (\S\ref{ssec:SandT}) that 
the group homomorphism 
\[
\TrFrob{G}: \CSiso{G} \to G(\Fq)^*
\]
provided by the trace of Frobenius is surjective with kernel $\Hh^2(\bG,\EEx)^{\Weil{}}$.
One can show, using the K\"unneth formula, that $\Hh^2(\bG,\EEx)$ is non-trivial unless $\bG$ is cyclic.
%The fact that $\TrFrob{G}$ may not be an isomorphism should not be too troubling: if quasicharacter sheaves $(\gcs{L}, \mu,\phi)$ and $(\gcs{L}', \mu',\phi')$ determine the same character of $G(\Fq)$ then $(\gcs{L},\phi)$ and $(\gcs{L}',\phi')$ are isomorphic as Weil sheaves.

@@

Having understood quasicharacter sheaves on commutative smooth group schemes over $\Fq$ in two extreme cases --
connected commutative algebraic groups and \'etale commutative group schemes -- we turn to the general case, using the component group sequence
\[
0 \to G^0 \to G \to \pi_0(G) \to 0.
\]
Using pullbacks of quasicharacter sheaves (\S\ref{ssec:Frob}), we obtain a diagram
\[
  \begin{tikzcd}[row sep=20, column sep=20]
    0 \rar & \CSiso{\pi_0(G)} \rar \dar{\TrFrob{\pi_0(G)}}
    & \CSiso{G} \rar \dar{\TrFrob{G}} & \CSiso{G^0} \rar \dar{\TrFrob{G^0}} & 0\\
    0 \rar & \pi_0(G)(\Fq)^* \rar & G(\Fq)^* \rar & G^0(\Fq)^* \rar & 0.
  \end{tikzcd}
\]
The rows of this diagram are exact (\S\S\ref{ssec:restriction},\ref{ssec:component}), so we may apply the snake lemma to prove the main theorem of the paper,
\begin{theorem*}[{Thm. \ref{thm:snake}}]
If $G$ is a smooth commutative group scheme over $\Fq$ then
\[
\TrFrob{G} : \CSiso{G} \to G(\Fq)^*
\]
is surjective with kernel $\Hh^0(\Weil{},\Hh^2(\pi_0(\bG),\EEx))$.
\end{theorem*}
\noindent
It follows that, unless the etale group scheme $\pi_0(\bG)$ is cyclic, the category $\CS(G)$ contains non-trivial character sheaves lying in the kernel of the trace of Frobenius and therefore lost in translation to characters of $G(\Fq)$ and, moreover, that these {\it invisible character sheaves} are parametrized by $\Hh^0(\Weil{},\Hh^2(\pi_0(\bG),\EEx))$.

@@

Throughout this paper, $G$ is a smooth commutative group scheme
over a finite field $\Fq$ and $m : \GxG\to G$ is its multiplication morphism.
Thus, the structure morphism $G \to \Spec{\Fq}$ is locally of finite type.
If the structure morphism $G \to \Spec{\Fq}$ is \'etale, then $G$ is an etale group scheme. 
If the structure morphism $G \to \Spec{\Fq}$ is of finite type, then $G$ is an algebraic group.
We will make use of the short exact sequence of smooth group schemes defining the component group scheme for $G$:
\[
\begin{tikzcd}
0 \rar & G^0 \arrow{r}{\iota_0} & G \arrow{r}{\pi_0} & \pi_0(G) \rar & 0.
\end{tikzcd}
\]
Then $G^0$ is a connected algebraic group and $\pi_0(G)$ is an \'etale commutative group scheme.
We remark that the smooth group scheme $G$ is an algebraic group if and only if the etale group scheme $\pi_0(G)$ is finite.

@@

\begin{proof}
Let $\pi: B \to G^0$ be a discrete isogeny, and set $A \ceq \ker \pi$.
  We will find a discrete isogeny $f: H\to G$
  such that that $H^0 = B$, $f^0 =\pi$ and
  $\pi_0(f) : \pi_0(H)\to \pi_0(G)$ is an isomorphism of component
  groups.  Namely, we will fit $\pi$ into the following diagram,
  \begin{equation}\label{extension-diagram}
  \begin{tikzcd}
  A \arrow[equal]{r} \dar & A \dar \\
  B \rar \dar[swap]{\pi} & H \rar \dar[swap]{f} & \pi_0(H) \arrow{d}[below,rotate=90]{\sim}[swap]{\pi_0(f)} \\
  G^0 \rar & G \rar & \pi_0(G),
  \end{tikzcd}
  \end{equation}
  where all rows and columns are exact and all maps are defined over
  $\Fq$.  We will do so by passing back and forth between group
  schemes over $\Fq$ and their $\bFq$-points.

>>>>

\begin{abstract}
We consider the category of character sheaves on smooth commutative group schemes $G$ over finite fields $k$ and expand the scope of the function-sheaf dictionary from connected commutative algebraic groups to this setting.
We show that the group of isomorphism classes of character sheaves is an extension of the group of characters of $G(k)$ by a cohomology group determined by the component group scheme of $G$.
As an application, we find a category of sheaves for which isomorphism classes correspond to quasi\-characters of algebraic tori, abelian varieties and unipotent wound groups over non-Archimedean local fields. 
In this geometrization of quasicharacters, our sheaves are local systems on
Greenberg transforms of locally finite type N\'eron models.
\end{abstract}

@@

%This paper extends Deligne's function-sheaf dictionary from connected commutative algebraic groups to smooth commutative group schemes over finite fields and then provides an application to the geometrization and categorification of quasicharacters of algebraic tori, abelian varieties and unipotent wound groups over non-Archimedean local fields.

@@

We begin the paper by returning to the case of connected commutative algebraic groups $G$ and thus revisiting Deligne's function-sheaf dictionary (\S\ref{sec:category}).
We consider character sheaves that arise via base change to $\bFq$ from local systems on $G$ (\S\ref{ssec:descentG}) and
those that appear in a pushforward from a constant sheaf along a discrete isogeny $H \to G$ (\S\ref{ssec:discrete_isogenies}).  
While these constructions make sense even for non-connected $G$, in the connected case we show that every character sheaf can be described in both of these ways (\S\ref{ssec:connected}).  
We use this fact to prove that $\TrFrob{G}$ is an isomorphism for connected commutative algebraic groups $G$
and to determine the automorphism groups of character sheaves on such $G$. 
These results are all well know.

Next, we consider character sheaves on \'etale commutative group schemes $G$ over $\Fq$ (\S\ref{sec:etale}).
\'Etale group schemes form a counterpoint to connected algebraic groups, since the component group of any smooth group scheme is an \'etale group scheme.
Our key tool for understanding the trace of Frobenius in the \'etale case 
%are a reinterpretation of $\CS(G)$ in terms of stalks (\S\ref{ssec:stalks}) and 
is a relationship with the Hochschild-Serre spectral sequence (\S\ref{ssec:E}) for $\Weil{} \ltimes \bG$, where $\Weil{} \subset \Gal(\Fq/\Fq)$ is the Weil group for $\Fq$.
We define (\S\ref{ssec:S}) an isomorphism $S_G$ from $\CSiso{G}$ to the second cohomology of the total space of the spectral sequence
\[
E_2^{p,q} \ceq \Hh^p(\Weil{}, \Hh^q(\bG, \EEx)) \Rightarrow \Hh^{p+q}(\Weil{} \ltimes \bG, \EEx).
\]
Paired with the short exact sequence
\[
  0 \to \Hh^0(\Weil{},\Hh^2({\bar G},\EEx)) \to \Hh^2(E^\bullet_G) \to \Hh^1(\Weil{},\Hh^1({\bar G},\EEx)) \to 0
\]
arising from the spectral sequence, the isomorphism $S_G$ allows us to show (\S\ref{ssec:SandT}) that 
the group homomorphism 
\[
\TrFrob{G}: \CSiso{G} \to G(\Fq)^*
\]
provided by the trace of Frobenius is surjective with kernel $\Hh^2(\bG,\EEx)^{\Weil{}}$.
One can show, using the K\"unneth formula, that $\Hh^2(\bG,\EEx)$ is non-trivial unless $\bG$ is cyclic.
%The fact that $\TrFrob{G}$ may not be an isomorphism should not be too troubling: if quasicharacter sheaves $(\gcs{L}, \mu,\phi)$ and $(\gcs{L}', \mu',\phi')$ determine the same character of $G(\Fq)$ then $(\gcs{L},\phi)$ and $(\gcs{L}',\phi')$ are isomorphic as Weil sheaves.

@@

Having understood character sheaves on commutative smooth group schemes over $\Fq$ in two extreme cases --
connected commutative algebraic groups and \'etale commutative group schemes -- we turn to the general case  (\S\ref{sec:main}), using the component group sequence
\[
0 \to G^0 \to G \to \pi_0(G) \to 0.
\]
Using pullbacks of character sheaves we obtain a diagram
\[
  \begin{tikzcd}[row sep=20, column sep=20]
    0 \rar & \CSiso{\pi_0(G)} \rar \dar{\TrFrob{\pi_0(G)}}
    & \CSiso{G} \rar \dar{\TrFrob{G}} & \CSiso{G^0} \rar \dar{\TrFrob{G^0}} & 0\\
    0 \rar & \pi_0(G)(\Fq)^* \rar & G(\Fq)^* \rar & G^0(\Fq)^* \rar & 0.
  \end{tikzcd}
\]
The rows of this diagram are exact (\S\S\ref{ssec:restriction},\ref{ssec:component}), so we may apply the snake lemma to prove the main theorem of the paper,
\begin{theorem*}[{Thm. \ref{thm:snake}}]
If $G$ is a smooth commutative group scheme over $\Fq$ then
\[
\TrFrob{G} : \CSiso{G} \to G(\Fq)^*
\]
is surjective with kernel $\Hh^0(\Weil{},\Hh^2(\pi_0(\bG),\EEx))$.
\end{theorem*}
\noindent
It follows that, unless the etale group scheme $\pi_0(\bG)$ is cyclic, the category $\CS(G)$ contains non-trivial character sheaves lying in the kernel of the trace of Frobenius and therefore lost in translation to characters of $G(\Fq)$ and, moreover, that these {\it invisible character sheaves} are parametrized by $\Hh^0(\Weil{},\Hh^2(\pi_0(\bG),\EEx))$.

@@

Throughout this paper, $G$ is a smooth commutative group scheme
over a finite field $\Fq$ and $m : G \times G\to G$ is its multiplication morphism.
Thus, the structure morphism $G \to \Spec{\Fq}$ is locally of finite type.
If the structure morphism $G \to \Spec{\Fq}$ is \'etale, then $G$ is an etale group scheme. 
If the structure morphism $G \to \Spec{\Fq}$ is of finite type, then $G$ is an algebraic group.
We will make use of the short exact sequence of smooth group schemes defining the component group scheme for $G$:
\[
\begin{tikzcd}
0 \rar & G^0 \arrow{r}{\iota_0} & G \arrow{r}{\pi_0} & \pi_0(G) \rar & 0.
\end{tikzcd}
\]
Then $G^0$ is a connected algebraic group and $\pi_0(G)$ is an \'etale commutative group scheme.
We remark that the smooth group scheme $G$ is an algebraic group if and only if the etale group scheme $\pi_0(G)$ is finite.

@@

\begin{proof}
Let $\pi: B \to G^0$ be a discrete isogeny, and set $A \ceq \ker \pi$.
  We will find a discrete isogeny $f: H\to G$
  such that that $H^0 = B$, $f^0 =\pi$ and
  $\pi_0(f) : \pi_0(H)\to \pi_0(G)$ is an isomorphism of component
  groups.  Namely, we will fit $\pi$ into the following diagram,
  \begin{equation}\label{extension-diagram}
  \begin{tikzcd}
  A \arrow{r} \dar & A \dar \\
  B \rar \dar[swap]{\pi} & H \rar \dar[swap]{f} & \pi_0(H) \arrow{d}[below,rotate=90]{\sim}[swap]{\pi_0(f)} \\
  G^0 \rar & G \rar & \pi_0(G),
  \end{tikzcd}
  \end{equation}
  where all rows and columns are exact and all maps are defined over
  $\Fq$.  We will do so by passing back and forth between group
  schemes over $\Fq$ and their $\bFq$-points.
